---
title: "16S Metagenomic Analysis - Enhanced Visualizations with Read Count Heatmaps"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 600
)
```

# Overview

This R Markdown document performs 16S rRNA gene amplicon sequencing visualization with:

- **Faceted plots** for before/after depletion comparisons
- **Full sample names** on x-axis
- **Complete relative abundance** (100% bars with single "Other" block at bottom)
- **Read count heatmap strip** at the bottom of each plot
- **Time-ordered enrichment plots** grouped by broth type
- **Publication-quality** figures with consistent color schemes

---

# Load Required Libraries

```{r load-libraries}
library(phyloseq)
library(dada2)
library(dplyr)
library(tidyr)
library(stringr)
library(readxl)
library(ggplot2)
library(RColorBrewer)
library(scales)
library(patchwork)
library(pheatmap)
library(cowplot)
library(rlang)

cat("✓ Packages loaded\n")
```

---

# Define Color Palettes

```{r define-color-palettes}
# Category colors for sample metadata
category_colors <- c(
  "Enrichment Testing" = "#e74c3c",
  "Host-DNA depletion" = "#3498db",
  "Filtration testing for Host-DNA depletion" = "#2ecc71",
  "Host-DNA depletion - enriched samples" = "#f39c12",
  "DNA extraction kit testing" = "#9b59b6",
  "Control" = "#95a5a6"
)

# Read count heatmap gradient (warm yellow-orange-red)
read_count_colors <- c("#fff7ec", "#fee8c8", "#fdd49e", "#fdbb84", 
                       "#fc8d59", "#ef6548", "#d7301f", "#990000")

# Heatmap colors for abundance (warm palette)
abundance_heatmap_colors <- c("white", "#ffeda0", "#fed976", "#feb24c", 
                               "#fd8d3c", "#fc4e2a", "#e31a1c", "#bd0026")

# ============================================================================
# CONSISTENT COLOR PALETTE SYSTEM
# ============================================================================
# Using ColorBrewer "Set3" palette for taxa
# Campylobacter manually assigned to green
# Other and Unclassified are GREY

# Fixed colors for special taxa
FIXED_TAXA_COLORS <- c(
  "Campylobacteraceae" = "#66C2A5",
  "Campylobacter" = "#66C2A5"
)

# Official ColorBrewer Set3 palette (12 colors)
# Removed the grey (#D9D9D9) to avoid confusion with Other/Unclassified
SET3_PALETTE <- c(
  "#A6CEE3",
  "#1F78B4",
  "#B2DF8A",
  "#33A02C",
  "#FB9A99",
  "#ff5d52",
  "#FDBF6F",
  "#FF7F00",
  "#CAB2D6",
  "#6A3D9A",
  "#B3CDE3",
  "#B15928"
)


# Function to get consistent color for a taxon
get_taxon_color <- function(taxon_name) {
  # Check if taxon has a fixed color (Campylobacter)
  if (taxon_name %in% names(FIXED_TAXA_COLORS)) {
    return(FIXED_TAXA_COLORS[[taxon_name]])
  }
  
  # For other taxa, assign from Set3 palette based on name hash
  hash_val <- (sum(utf8ToInt(taxon_name)) %% length(SET3_PALETTE)) + 1
  return(SET3_PALETTE[hash_val])
}

# Helper function to create color palette with consistent colors
create_color_palette <- function(top_taxa) {
  # Get colors for each taxon
  taxa_colors <- sapply(top_taxa, get_taxon_color)
  
  # Check for duplicate colors and reassign if needed
  used_colors <- c()
  for (i in seq_along(taxa_colors)) {
    if (taxa_colors[i] %in% used_colors) {
      available <- setdiff(SET3_PALETTE, used_colors)
      if (length(available) > 0) {
        taxa_colors[i] <- available[1]
      }
    }
    used_colors <- c(used_colors, taxa_colors[i])
  }
  
  # Order: Other (light grey), Unclassified (medium grey), then top taxa
  all_cols <- c(
    "Other" = "#C0C0C0",
    "Unclassified" = "#808080",
    taxa_colors
  )
  
  return(all_cols)
}

cat("✓ Color palettes defined\n")
```

---

# Setup Sample Data

```{r setup-sample-data}
cat("Current phyloseq object:\n")
print(ps)

# Load DADA2 track data
track_df <- as.data.frame(track)
track_df$Sample <- rownames(track_df)

# Load sample metadata
sample_metadata <- read_excel(
  "/nesi/nobackup/massey04083/method_development_chapter1/16S_Sequencing/16S samples - sample descriptions.xlsx"
)

# Merge and prepare metadata
merged_data <- track_df %>%
  left_join(sample_metadata, by = c("Sample" = "Sample code")) %>%
  mutate(
    Method_Category = `Method development section`,
    Full_Name = `Full sample name`,
    DNA_Conc = as.numeric(gsub("Too Low", NA, gsub(" ", "", `Concentration (ng/ul)`))),
    input_reads = input,
    filtered_reads = filtered,
    merged_reads = merged,
    nonchim_reads = nonchim
  )

# Align with phyloseq sample order
rownames(merged_data) <- merged_data$Sample
merged_data <- merged_data[sample_names(ps), , drop = FALSE]
sample_data(ps) <- sample_data(merged_data)

cat("✓ Sample data integrated\n")
```

---

# Fix Metadata Issues

```{r fix-metadata-REQUIRED}
cat("=== FIXING METADATA ISSUES ===\n\n")

# Remove Undetermined samples
if ("Undetermined" %in% sample_names(ps)) {
  ps <- prune_samples(sample_names(ps) != "Undetermined", ps)
  cat("✓ Removed 'Undetermined' sample\n")
}

samp_df <- data.frame(sample_data(ps), stringsAsFactors = FALSE, check.names = FALSE)

# Fix QDneasy samples
qd_samples <- c("QDnBT1", "QDnBT2", "QDnBT3")
for (i in seq_along(qd_samples)) {
  if (qd_samples[i] %in% rownames(samp_df)) {
    samp_df[qd_samples[i], "Method_Category"] <- "DNA extraction kit testing"
    if ("Full_Name" %in% colnames(samp_df)) {
      samp_df[qd_samples[i], "Full_Name"] <- paste0("QDneasy Blood Tissue Kit Rep", i)
    }
  }
}

# Create display names
samp_df$Full_Name_Display <- NA_character_

for (i in seq_len(nrow(samp_df))) {
  sample_code <- rownames(samp_df)[i]
  full_name <- if ("Full_Name" %in% colnames(samp_df)) samp_df[i, "Full_Name"] else NA

  if (is.na(full_name) || full_name == "") {
    samp_df$Full_Name_Display[i] <- sample_code
  } else {
    samp_df$Full_Name_Display[i] <- paste0(full_name, " [", sample_code, "]")
  }
}

sample_data(ps) <- sample_data(samp_df)

cat("\n✓ Verification:\n")
cat("  Missing Method_Category:", sum(is.na(samp_df$Method_Category)), "\n")
cat("  Missing nonchim_reads:", sum(is.na(samp_df$nonchim_reads)), "\n")
cat("  Duplicate Full_Name_Display:", sum(duplicated(samp_df$Full_Name_Display)), "\n")
cat("\nSample categories:\n")
print(table(sample_data(ps)$Method_Category, useNA = "ifany"))
cat("\n✓ Metadata fix complete!\n")
```

---

# Add Enrichment Metadata and Taxonomy Aggregation

```{r enrichment-and-tax-setup}
samp_df <- data.frame(sample_data(ps), stringsAsFactors = FALSE, check.names = FALSE)

# Extract broth type
samp_df$Broth_Type <- dplyr::case_when(
  str_detect(samp_df$Full_Name_Display, regex("CAT", ignore_case = TRUE)) &
    str_detect(samp_df$Full_Name_Display, regex("Bolton|BFBB|BF\\s*Bolton", ignore_case = TRUE)) ~ "BF Bolton + CAT",
  str_detect(samp_df$Full_Name_Display, regex("Bolton|BFBB|BF\\s*Bolton", ignore_case = TRUE)) ~ "BF Bolton",
  str_detect(samp_df$Full_Name_Display, regex("\\bRVS\\b", ignore_case = TRUE)) ~ "RVS",
  str_detect(samp_df$Full_Name_Display, regex("\\bBPW\\b", ignore_case = TRUE)) ~ "BPW",
  TRUE ~ NA_character_
)

# Extract incubation time
samp_df$Incubation_hr <- as.numeric(str_extract(samp_df$Full_Name_Display, "(?i)\\d+(?=\\s*h)"))

sample_data(ps) <- sample_data(samp_df)

cat("✓ Added Broth_Type + Incubation_hr\n")
print(table(sample_data(ps)$Broth_Type, useNA = "ifany"))
print(summary(sample_data(ps)$Incubation_hr))

# Taxonomic aggregation (NArm=FALSE keeps Unclassified taxa)
ps_family <- tax_glom(ps, taxrank = "Family", NArm = FALSE)
ps_genus  <- tax_glom(ps, taxrank = "Genus",  NArm = FALSE)

# ENFORCED compositional transformation (guarantees sum = 1 per sample)
ps_family_rel <- transform_sample_counts(ps_family, function(x) {
  total <- sum(x, na.rm = TRUE)
  if (total > 0) x / total else x
})

ps_genus_rel <- transform_sample_counts(ps_genus, function(x) {
  total <- sum(x, na.rm = TRUE)
  if (total > 0) x / total else x
})

cat("✓ Taxonomic aggregations complete (Unclassified retained)\n")

# Verify closure
cat("\nPost-transformation sample sums (Family):\n")
print(summary(sample_sums(ps_family_rel)))
cat("\nPost-transformation sample sums (Genus):\n")
print(summary(sample_sums(ps_genus_rel)))
```

---

# Define Plotting Functions

```{r define-plotting-functions}
# ============================================================
# PLOTTING FUNCTIONS (CLEAN + FIXED + ADDED NO-FILTER PANEL)
# - guarantees bars sum to 100% via renormalization
# - preserves real "Other" (no remainder/diff hacks)
# - NEW: create_triple_faceted_plot for unfiltered + filtered views
# ============================================================

normalize_composition_by_sample <- function(df, sample_col = "Sample", abundance_col = "Abundance") {
  sample_sym <- rlang::sym(sample_col)
  abund_sym  <- rlang::sym(abundance_col)

  df %>%
    dplyr::mutate(!!abund_sym := pmax(0, !!abund_sym)) %>%
    dplyr::group_by(!!sample_sym) %>%
    dplyr::mutate(
      .total = sum(!!abund_sym, na.rm = TRUE),
      !!abund_sym := dplyr::if_else(.total > 0, (!!abund_sym) / .total, 0)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(-.total)
}

# NEW FUNCTION: Three-panel plot (All samples + Host-Depleted + No Host-Depletion)
create_triple_faceted_plot <- function(physeq,
                                       sample_filter = NULL,
                                       title = "Microbial Composition - Complete Overview",
                                       output_file = NULL,
                                       top_n = 12,
                                       taxon_rank = "Family",
                                       width = 20,
                                       height = 11) {

  if (!is.null(sample_filter) && "Method_Category" %in% colnames(sample_data(physeq))) {
    physeq <- prune_samples(sample_data(physeq)$Method_Category %in% sample_filter, physeq)
  }
  if (nsamples(physeq) == 0) {
    warning("No samples found after filtering")
    return(NULL)
  }

  plot_df <- psmelt(physeq)
  if ("sample_Sample" %in% colnames(plot_df) && !"Sample" %in% colnames(plot_df)) {
    plot_df$Sample <- plot_df$sample_Sample
  }

  plot_df$Taxon <- plot_df[[taxon_rank]]
  plot_df$Taxon[is.na(plot_df$Taxon) | plot_df$Taxon == ""] <- "Unclassified"

  top_taxa <- plot_df %>%
    dplyr::filter(Taxon != "Unclassified") %>%
    dplyr::group_by(Taxon) %>%
    dplyr::summarise(Total = sum(Abundance), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(Total)) %>%
    dplyr::slice_head(n = top_n) %>%
    dplyr::pull(Taxon)

  plot_df <- plot_df %>%
    dplyr::mutate(Taxon_Display = dplyr::case_when(
      Taxon == "Unclassified" ~ "Unclassified",
      Taxon %in% top_taxa ~ Taxon,
      TRUE ~ "Other"
    ))

  plot_df_agg <- plot_df %>%
    dplyr::group_by(Sample, Taxon_Display) %>%
    dplyr::summarise(
      Abundance = sum(Abundance, na.rm = TRUE),
      Full_Name_Display = dplyr::first(Full_Name_Display),
      Method_Category   = dplyr::first(Method_Category),
      nonchim_reads     = dplyr::first(nonchim_reads),
      .groups = "drop"
    )

  plot_df_agg <- normalize_composition_by_sample(plot_df_agg, sample_col = "Sample", abundance_col = "Abundance")
  
  # Safety: ensure no NA/negative values before plotting
  plot_df_agg <- plot_df_agg %>%
    dplyr::filter(!is.na(Abundance), Abundance >= 0) %>%
    dplyr::mutate(Abundance = pmax(Abundance, 1e-10))

  # Create depletion status
  plot_df_agg$Depletion_Status <- ifelse(
    grepl("No host|No Host", plot_df_agg$Full_Name_Display, ignore.case = TRUE),
    "No Host-Depletion",
    "With Host-Depletion"
  )

  # Create three datasets: All, No Host-Depletion, With Host-Depletion
  plot_df_all <- plot_df_agg %>%
    dplyr::mutate(Panel = "All Samples")
  
  plot_df_no_host <- plot_df_agg %>%
    dplyr::filter(Depletion_Status == "No Host-Depletion") %>%
    dplyr::mutate(Panel = "No Host-Depletion")
  
  plot_df_with_host <- plot_df_agg %>%
    dplyr::filter(Depletion_Status == "With Host-Depletion") %>%
    dplyr::mutate(Panel = "With Host-Depletion")
  
  # Combine all three
  plot_df_combined <- dplyr::bind_rows(plot_df_all, plot_df_no_host, plot_df_with_host)
  
  # Set panel order
  plot_df_combined$Panel <- factor(
    plot_df_combined$Panel, 
    levels = c("All Samples", "No Host-Depletion", "With Host-Depletion")
  )

  plot_df_combined$Sample_Label <- plot_df_combined$Full_Name_Display

  # Order samples within each panel
  plot_df_combined <- plot_df_combined %>%
    dplyr::arrange(Panel, Depletion_Status, Sample) %>%
    dplyr::group_by(Panel) %>%
    dplyr::mutate(Sample_Label = factor(Sample_Label, levels = unique(Sample_Label))) %>%
    dplyr::ungroup()

  taxa_order <- c("Other", "Unclassified", top_taxa)
  plot_df_combined$Taxon_Display <- factor(plot_df_combined$Taxon_Display, levels = taxa_order)

  cols <- create_color_palette(top_taxa)

  p_bar <- ggplot(plot_df_combined, aes(x = Sample_Label, y = Abundance * 100, fill = Taxon_Display)) +
    geom_bar(stat = "identity", color = "white", linewidth = 0.2) +
    scale_fill_manual(values = cols, name = taxon_rank, drop = FALSE) +
    facet_wrap(~Panel, scales = "free_x", ncol = 3) +
    labs(y = "Relative Abundance (%)", x = "", title = title) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      strip.background = element_rect(fill = "grey90", color = "black"),
      strip.text = element_text(face = "bold", size = 12),
      panel.grid.major.y = element_line(color = "grey90"),
      legend.position = "right",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.margin = margin(5, 5, 0, 5)
    ) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 100.1))

  read_df <- plot_df_combined %>% dplyr::distinct(Sample_Label, Panel, nonchim_reads)
  max_reads <- max(read_df$nonchim_reads, na.rm = TRUE)

  p_heat <- ggplot(read_df, aes(x = Sample_Label, y = 1, fill = nonchim_reads)) +
    geom_tile(color = "white", linewidth = 0.5) +
    scale_fill_gradientn(
      colors = read_count_colors,
      name = "Read Count",
      labels = scales::comma,
      breaks = c(0, max_reads),
      limits = c(0, max_reads)
    ) +
    facet_wrap(~Panel, scales = "free_x", ncol = 3) +
    theme_void() +
    theme(
      strip.text = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
      legend.position = "right",
      legend.direction = "vertical",
      legend.key.width = unit(0.4, "cm"),
      legend.key.height = unit(1.5, "cm"),
      plot.margin = margin(0, 5, 5, 5)
    ) +
    scale_x_discrete(position = "bottom")

  p_out <- (p_bar / p_heat) +
    patchwork::plot_layout(heights = c(10, 1.5)) +
    patchwork::plot_annotation(theme = theme(plot.margin = margin(5, 5, 5, 5)))

  if (!is.null(output_file)) {
    if (!dir.exists(dirname(output_file))) dir.create(dirname(output_file), recursive = TRUE)
    png(output_file, width = width, height = height, units = "in", res = 300)
    print(p_out)
    dev.off()
    cat("✓ Saved:", output_file, "\n")
  }

  return(p_out)
}

# ORIGINAL create_faceted_plot (unchanged)
create_faceted_plot <- function(physeq,
                                sample_filter = NULL,
                                title = "Microbial Composition",
                                output_file = NULL,
                                top_n = 12,
                                taxon_rank = "Family",
                                width = 16,
                                height = 11) {

  if (!is.null(sample_filter) && "Method_Category" %in% colnames(sample_data(physeq))) {
    physeq <- prune_samples(sample_data(physeq)$Method_Category %in% sample_filter, physeq)
  }
  if (nsamples(physeq) == 0) {
    warning("No samples found after filtering")
    return(NULL)
  }

  plot_df <- psmelt(physeq)
  if ("sample_Sample" %in% colnames(plot_df) && !"Sample" %in% colnames(plot_df)) {
    plot_df$Sample <- plot_df$sample_Sample
  }

  plot_df$Taxon <- plot_df[[taxon_rank]]
  plot_df$Taxon[is.na(plot_df$Taxon) | plot_df$Taxon == ""] <- "Unclassified"

  top_taxa <- plot_df %>%
    dplyr::filter(Taxon != "Unclassified") %>%
    dplyr::group_by(Taxon) %>%
    dplyr::summarise(Total = sum(Abundance), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(Total)) %>%
    dplyr::slice_head(n = top_n) %>%
    dplyr::pull(Taxon)

  plot_df <- plot_df %>%
    dplyr::mutate(Taxon_Display = dplyr::case_when(
      Taxon == "Unclassified" ~ "Unclassified",
      Taxon %in% top_taxa ~ Taxon,
      TRUE ~ "Other"
    ))

  plot_df_agg <- plot_df %>%
    dplyr::group_by(Sample, Taxon_Display) %>%
    dplyr::summarise(
      Abundance = sum(Abundance, na.rm = TRUE),
      Full_Name_Display = dplyr::first(Full_Name_Display),
      Method_Category   = dplyr::first(Method_Category),
      nonchim_reads     = dplyr::first(nonchim_reads),
      .groups = "drop"
    )

  plot_df_agg <- normalize_composition_by_sample(plot_df_agg, sample_col = "Sample", abundance_col = "Abundance")
  
  # Safety: ensure no NA/negative values before plotting
  plot_df_agg <- plot_df_agg %>%
    dplyr::filter(!is.na(Abundance), Abundance >= 0) %>%
    dplyr::mutate(Abundance = pmax(Abundance, 1e-10))

  plot_df_agg$Depletion <- ifelse(
    grepl("No host|No Host", plot_df_agg$Full_Name_Display, ignore.case = TRUE),
    "No Host-Depletion",
    "With Host-Depletion"
  )

  plot_df_agg$Sample_Label <- plot_df_agg$Full_Name_Display

  plot_df_agg <- plot_df_agg %>%
    dplyr::arrange(Depletion, Sample) %>%
    dplyr::group_by(Depletion) %>%
    dplyr::mutate(Sample_Label = factor(Sample_Label, levels = unique(Sample_Label))) %>%
    dplyr::ungroup()

  taxa_order <- c("Other", "Unclassified", top_taxa)
  plot_df_agg$Taxon_Display <- factor(plot_df_agg$Taxon_Display, levels = taxa_order)

  cols <- create_color_palette(top_taxa)

  p_bar <- ggplot(plot_df_agg, aes(x = Sample_Label, y = Abundance * 100, fill = Taxon_Display)) +
    geom_bar(stat = "identity", color = "white", linewidth = 0.2) +
    scale_fill_manual(values = cols, name = taxon_rank, drop = FALSE) +
    facet_wrap(~Depletion, scales = "free_x", ncol = 2) +
    labs(y = "Relative Abundance (%)", x = "", title = title) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      strip.background = element_rect(fill = "grey90", color = "black"),
      strip.text = element_text(face = "bold", size = 12),
      panel.grid.major.y = element_line(color = "grey90"),
      legend.position = "right",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.margin = margin(5, 5, 0, 5)
    ) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 100.1))

  read_df <- plot_df_agg %>% dplyr::distinct(Sample_Label, Depletion, nonchim_reads)
  max_reads <- max(read_df$nonchim_reads, na.rm = TRUE)

  p_heat <- ggplot(read_df, aes(x = Sample_Label, y = 1, fill = nonchim_reads)) +
    geom_tile(color = "white", linewidth = 0.5) +
    scale_fill_gradientn(
      colors = read_count_colors,
      name = "Read Count",
      labels = scales::comma,
      breaks = c(0, max_reads),
      limits = c(0, max_reads)
    ) +
    facet_wrap(~Depletion, scales = "free_x", ncol = 2) +
    theme_void() +
    theme(
      strip.text = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
      legend.position = "right",
      legend.direction = "vertical",
      legend.key.width = unit(0.4, "cm"),
      legend.key.height = unit(1.5, "cm"),
      plot.margin = margin(0, 5, 5, 5)
    ) +
    scale_x_discrete(position = "bottom")

  p_out <- (p_bar / p_heat) +
    patchwork::plot_layout(heights = c(10, 1.5)) +
    patchwork::plot_annotation(theme = theme(plot.margin = margin(5, 5, 5, 5)))

  if (!is.null(output_file)) {
    if (!dir.exists(dirname(output_file))) dir.create(dirname(output_file), recursive = TRUE)
    png(output_file, width = width, height = height, units = "in", res = 300)
    print(p_out)
    dev.off()
    cat("✓ Saved:", output_file, "\n")
  }

  return(p_out)
}

# (Rest of the functions remain unchanged: create_simple_plot, create_enrichment_plot_4broths, create_heatmap)

create_simple_plot <- function(physeq,
                               sample_filter = NULL,
                               title = "Microbial Composition",
                               output_file = NULL,
                               top_n = 12,
                               taxon_rank = "Family",
                               width = 14,
                               height = 9) {

  if (!is.null(sample_filter) && "Method_Category" %in% colnames(sample_data(physeq))) {
    physeq <- prune_samples(sample_data(physeq)$Method_Category %in% sample_filter, physeq)
  }
  if (nsamples(physeq) == 0) {
    warning("No samples found after filtering")
    return(NULL)
  }

  plot_df <- psmelt(physeq)
  if ("sample_Sample" %in% colnames(plot_df) && !"Sample" %in% colnames(plot_df)) {
    plot_df$Sample <- plot_df$sample_Sample
  }

  plot_df$Taxon <- plot_df[[taxon_rank]]
  plot_df$Taxon[is.na(plot_df$Taxon) | plot_df$Taxon == ""] <- "Unclassified"

  top_taxa <- plot_df %>%
    dplyr::filter(Taxon != "Unclassified") %>%
    dplyr::group_by(Taxon) %>%
    dplyr::summarise(Total = sum(Abundance), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(Total)) %>%
    dplyr::slice_head(n = top_n) %>%
    dplyr::pull(Taxon)

  plot_df <- plot_df %>%
    dplyr::mutate(Taxon_Display = dplyr::case_when(
      Taxon == "Unclassified" ~ "Unclassified",
      Taxon %in% top_taxa ~ Taxon,
      TRUE ~ "Other"
    ))

  plot_df_agg <- plot_df %>%
    dplyr::group_by(Sample, Taxon_Display) %>%
    dplyr::summarise(
      Abundance = sum(Abundance, na.rm = TRUE),
      Full_Name_Display = dplyr::first(Full_Name_Display),
      Method_Category   = dplyr::first(Method_Category),
      nonchim_reads     = dplyr::first(nonchim_reads),
      .groups = "drop"
    )

  plot_df_agg <- normalize_composition_by_sample(plot_df_agg, sample_col = "Sample", abundance_col = "Abundance")
  
  # Safety: ensure no NA/negative values before plotting
  plot_df_agg <- plot_df_agg %>%
    dplyr::filter(!is.na(Abundance), Abundance >= 0) %>%
    dplyr::mutate(Abundance = pmax(Abundance, 1e-10))

  plot_df_agg$Sample_Label <- plot_df_agg$Full_Name_Display
  plot_df_agg <- plot_df_agg %>%
    dplyr::mutate(Sample_Label = factor(Sample_Label, levels = unique(Sample_Label)))

  taxa_order <- c("Other", "Unclassified", top_taxa)
  plot_df_agg$Taxon_Display <- factor(plot_df_agg$Taxon_Display, levels = taxa_order)

  cols <- create_color_palette(top_taxa)

  p_bar <- ggplot(plot_df_agg, aes(x = Sample_Label, y = Abundance * 100, fill = Taxon_Display)) +
    geom_bar(stat = "identity", color = "white", linewidth = 0.2) +
    scale_fill_manual(values = cols, name = taxon_rank, drop = FALSE) +
    labs(y = "Relative Abundance (%)", x = "", title = title) +
    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      panel.grid.major.y = element_line(color = "grey90"),
      legend.position = "right",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.margin = margin(5, 5, 0, 5)
    ) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 100.1))

  read_df <- plot_df_agg %>% dplyr::distinct(Sample_Label, nonchim_reads)
  max_reads <- max(read_df$nonchim_reads, na.rm = TRUE)

  p_heat <- ggplot(read_df, aes(x = Sample_Label, y = 1, fill = nonchim_reads)) +
    geom_tile(color = "white", linewidth = 0.5) +
    scale_fill_gradientn(
      colors = read_count_colors,
      name = "Read Count",
      labels = scales::comma,
      breaks = c(0, max_reads),
      limits = c(0, max_reads)
    ) +
    theme_void() +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
      legend.position = "right",
      legend.direction = "vertical",
      legend.key.width = unit(0.4, "cm"),
      legend.key.height = unit(1.5, "cm"),
      plot.margin = margin(0, 5, 5, 5)
    ) +
    scale_x_discrete(position = "bottom")

  p_out <- (p_bar / p_heat) +
    patchwork::plot_layout(heights = c(10, 1.5)) +
    patchwork::plot_annotation(theme = theme(plot.margin = margin(5, 5, 5, 5)))

  if (!is.null(output_file)) {
    if (!dir.exists(dirname(output_file))) dir.create(dirname(output_file), recursive = TRUE)
    png(output_file, width = width, height = height, units = "in", res = 300)
    print(p_out)
    dev.off()
    cat("✓ Saved:", output_file, "\n")
  }

  return(p_out)
}

create_enrichment_plot_4broths <- function(physeq,
                                           sample_filter = "Enrichment Testing",
                                           title = "Enrichment Testing - Temporal Dynamics by Broth",
                                           output_file = NULL,
                                           top_n = 12,
                                           taxon_rank = "Family",
                                           width = 16,
                                           height = 12) {

  if (!is.null(sample_filter) && "Method_Category" %in% colnames(sample_data(physeq))) {
    physeq <- prune_samples(sample_data(physeq)$Method_Category %in% sample_filter, physeq)
  }
  if (nsamples(physeq) == 0) {
    warning("No samples found after filtering")
    return(NULL)
  }

  plot_df <- psmelt(physeq)
  if ("sample_Sample" %in% colnames(plot_df) && !"Sample" %in% colnames(plot_df)) {
    plot_df$Sample <- plot_df$sample_Sample
  }

  req <- c("Broth_Type", "Incubation_hr", "nonchim_reads", "Full_Name_Display")
  miss <- req[!req %in% colnames(plot_df)]
  if (length(miss) > 0) {
    stop("Missing required metadata columns: ", paste(miss, collapse = ", "))
  }

  broth_levels <- c("RVS", "BPW", "BF Bolton", "BF Bolton + CAT")

  plot_df <- plot_df %>%
    dplyr::filter(!is.na(Broth_Type), !is.na(Incubation_hr)) %>%
    dplyr::filter(Broth_Type %in% broth_levels) %>%
    dplyr::mutate(Broth_Type = factor(Broth_Type, levels = broth_levels))

  plot_df$Taxon <- plot_df[[taxon_rank]]
  plot_df$Taxon[is.na(plot_df$Taxon) | plot_df$Taxon == ""] <- "Unclassified"

  top_taxa <- plot_df %>%
    dplyr::filter(Taxon != "Unclassified") %>%
    dplyr::group_by(Taxon) %>%
    dplyr::summarise(Total = sum(Abundance), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(Total)) %>%
    dplyr::slice_head(n = top_n) %>%
    dplyr::pull(Taxon)

  plot_df <- plot_df %>%
    dplyr::mutate(Taxon_Display = dplyr::case_when(
      Taxon == "Unclassified" ~ "Unclassified",
      Taxon %in% top_taxa ~ Taxon,
      TRUE ~ "Other"
    ))

  plot_df_agg <- plot_df %>%
    dplyr::group_by(Broth_Type, Incubation_hr, Taxon_Display) %>%
    dplyr::summarise(
      Abundance = sum(Abundance),
      .groups = "drop"
    )

  time_levels <- sort(unique(plot_df_agg$Incubation_hr))
  taxa_order <- c("Other", "Unclassified", top_taxa)

  plot_df_agg <- plot_df_agg %>%
    dplyr::mutate(
      Incubation_hr = factor(Incubation_hr, levels = time_levels),
      Taxon_Display = factor(Taxon_Display, levels = taxa_order)
    )

  read_df <- plot_df %>%
    dplyr::distinct(Sample, Broth_Type, Incubation_hr, nonchim_reads) %>%
    dplyr::group_by(Broth_Type, Incubation_hr) %>%
    dplyr::summarise(
      nonchim_reads = median(nonchim_reads, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      Incubation_hr = factor(Incubation_hr, levels = time_levels)
    )

  cols <- create_color_palette(top_taxa)

  max_reads_all <- max(read_df$nonchim_reads, na.rm = TRUE)

  create_single_broth_panel <- function(broth_name, show_y_axis = TRUE) {

    broth_data <- plot_df_agg %>% dplyr::filter(Broth_Type == broth_name)
    broth_reads <- read_df %>% dplyr::filter(Broth_Type == broth_name)

    if (nrow(broth_data) == 0) return(NULL)

    p_bar <- ggplot(broth_data, aes(x = Incubation_hr, y = Abundance * 100, fill = Taxon_Display)) +
      geom_bar(stat = "identity", color = "white", linewidth = 0.2, position = "stack") +
      scale_fill_manual(values = cols, name = taxon_rank, drop = FALSE) +
      labs(x = NULL, y = if(show_y_axis) "Relative Abundance (%)" else NULL,
           title = broth_name) +
      theme_classic(base_size = 11) +
      theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_text(size = 10),
        panel.grid.major.y = element_line(color = "grey90"),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
        plot.margin = margin(5, 5, 0, 5)
      ) +
      scale_y_continuous(expand = c(0, 0), limits = c(0, 100.5))

    if (!show_y_axis) {
      p_bar <- p_bar + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
    }

    p_heat <- ggplot(broth_reads, aes(x = Incubation_hr, y = 1, fill = nonchim_reads)) +
      geom_tile(color = "white", linewidth = 0.5) +
      scale_fill_gradientn(
        colors = read_count_colors,
        name = "Read Count",
        labels = scales::comma,
        breaks = c(0, max_reads_all),
        limits = c(0, max_reads_all)
      ) +
      labs(x = "Incubation (h)") +
      theme_void() +
      theme(
        axis.text.x = element_text(size = 9, margin = margin(t = 2)),
        axis.title.x = element_text(size = 9, margin = margin(t = 3)),
        legend.position = "none",
        plot.margin = margin(0, 5, 5, 5)
      )

    p_combined <- p_bar / p_heat + patchwork::plot_layout(heights = c(10, 1))

    return(p_combined)
  }

  p_rvs <- create_single_broth_panel("RVS", show_y_axis = TRUE)
  p_bpw <- create_single_broth_panel("BPW", show_y_axis = FALSE)
  p_bolton <- create_single_broth_panel("BF Bolton", show_y_axis = TRUE)
  p_bolton_cat <- create_single_broth_panel("BF Bolton + CAT", show_y_axis = FALSE)

  p_legend_bar <- ggplot(plot_df_agg, aes(x = Incubation_hr, y = Abundance * 100, fill = Taxon_Display)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = cols, name = taxon_rank, drop = FALSE) +
    theme(legend.position = "right",
          legend.title = element_text(face = "bold"),
          legend.key.size = unit(0.4, "cm"))

  legend_bar <- cowplot::get_legend(p_legend_bar)

  p_legend_heat <- ggplot(read_df, aes(x = Incubation_hr, y = 1, fill = nonchim_reads)) +
    geom_tile() +
    scale_fill_gradientn(
      colors = read_count_colors,
      name = "Read Count",
      labels = scales::comma,
      breaks = c(0, max_reads_all),
      limits = c(0, max_reads_all)
    ) +
    theme(legend.position = "right",
          legend.direction = "vertical",
          legend.key.width = unit(0.4, "cm"),
          legend.key.height = unit(2, "cm"),
          legend.title = element_text(face = "bold"))

  legend_heat <- cowplot::get_legend(p_legend_heat)

  panels <- list(p_rvs, p_bpw, p_bolton, p_bolton_cat)
  panels <- panels[!sapply(panels, is.null)]

  if (length(panels) == 4) {
    top_row <- (p_rvs | p_bpw)
    bottom_row <- (p_bolton | p_bolton_cat)
    main_plot <- top_row / bottom_row
  } else {
    main_plot <- patchwork::wrap_plots(panels, ncol = 2)
  }

  title_grob <- cowplot::ggdraw() +
    cowplot::draw_label(title, fontface = "bold", size = 14, hjust = 0.5)

  legends_combined <- cowplot::plot_grid(
    legend_bar,
    legend_heat,
    ncol = 1,
    rel_heights = c(1, 0.5),
    align = "v"
  )

  p_out <- cowplot::plot_grid(
    title_grob,
    cowplot::plot_grid(
      main_plot,
      legends_combined,
      ncol = 2,
      rel_widths = c(4, 1)
    ),
    ncol = 1,
    rel_heights = c(0.05, 1)
  )

  if (!is.null(output_file)) {
    if (!dir.exists(dirname(output_file))) dir.create(dirname(output_file), recursive = TRUE)
    png(output_file, width = width, height = height, units = "in", res = 300)
    print(p_out)
    dev.off()
    cat("✓ Saved:", output_file, "\n")
  }

  return(p_out)
}

create_heatmap <- function(physeq,
                           sample_filter = NULL,
                           title = "Microbial Diversity Heatmap",
                           output_file = NULL,
                           top_n = 12,
                           taxon_rank = "Family",
                           width = 20,
                           height = 12) {

  if (!is.null(sample_filter) && "Method_Category" %in% colnames(sample_data(physeq))) {
    physeq <- prune_samples(sample_data(physeq)$Method_Category %in% sample_filter, physeq)
  }
  if (nsamples(physeq) == 0) {
    warning("No samples found after filtering")
    return(NULL)
  }

  otu_mat <- as.matrix(otu_table(physeq))
  if (!taxa_are_rows(physeq)) otu_mat <- t(otu_mat)

  tax_df <- as.data.frame(tax_table(physeq))
  stopifnot(taxon_rank %in% colnames(tax_df))

  rownames(otu_mat) <- tax_df[[taxon_rank]]
  rownames(otu_mat)[is.na(rownames(otu_mat)) | rownames(otu_mat) == ""] <- "Unclassified"
  otu_mat <- rowsum(otu_mat, rownames(otu_mat))

  top_taxa <- names(sort(rowSums(otu_mat), decreasing = TRUE)[1:top_n])
  otu_mat_top <- otu_mat[top_taxa, , drop = FALSE] * 100

  meta_df <- data.frame(sample_data(physeq), check.names = FALSE, stringsAsFactors = FALSE)
  valid_samples <- rownames(meta_df)[!is.na(meta_df$Method_Category) & !is.na(meta_df$nonchim_reads)]

  if (length(valid_samples) == 0) {
    warning("No samples with complete metadata found")
    return(NULL)
  }

  meta_df <- meta_df[valid_samples, , drop = FALSE]
  otu_mat_top <- otu_mat_top[, valid_samples, drop = FALSE]

  sample_order <- rownames(meta_df)[order(meta_df$Method_Category)]
  otu_mat_top <- otu_mat_top[, sample_order, drop = FALSE]
  meta_df <- meta_df[sample_order, , drop = FALSE]

  display_names <- if ("Full_Name_Display" %in% colnames(meta_df) && !all(is.na(meta_df$Full_Name_Display))) {
    meta_df$Full_Name_Display
  } else {
    rownames(meta_df)
  }
  display_names <- ifelse(nchar(display_names) > 50, paste0(substr(display_names, 1, 47), "..."), display_names)
  if (any(duplicated(display_names))) display_names <- make.unique(display_names, sep = "_")

  colnames(otu_mat_top) <- display_names

  annotation_col <- data.frame(
    Category = meta_df$Method_Category,
    Reads = meta_df$nonchim_reads,
    row.names = display_names,
    stringsAsFactors = FALSE
  )

  ann_colors <- list(
    Category = category_colors[names(category_colors) %in% unique(annotation_col$Category)],
    Reads = colorRampPalette(c("white", "#fee391", "#fec44f", "#fe9929", "#d95f0e", "#993404"))(100)
  )

  if (!is.null(output_file)) {
    if (!dir.exists(dirname(output_file))) dir.create(dirname(output_file), recursive = TRUE)
    png(output_file, width = width, height = height, units = "in", res = 300)
  }

  tryCatch({
    pheatmap(
      otu_mat_top,
      color = colorRampPalette(abundance_heatmap_colors)(100),
      cluster_rows = FALSE,
      cluster_cols = FALSE,
      annotation_col = annotation_col,
      annotation_colors = ann_colors,
      show_colnames = TRUE,
      show_rownames = TRUE,
      fontsize = 9,
      fontsize_col = 7,
      fontsize_row = 10,
      angle_col = 90,
      main = title,
      border_color = "white",
      cellwidth = 12,
      cellheight = 12
    )
  }, error = function(e) {
    cat("Error creating heatmap:", e$message, "\n")
  })

  if (!is.null(output_file)) {
    dev.off()
    cat("✓ Saved:", output_file, "\n")
  }

  invisible(NULL)
}

cat("✓ All plotting functions defined\n")
```

---

# Diagnostic Check (optional - can comment out after verification)

```{r diagnostic-check-composition}
# Check if ps_family_rel samples sum to 1
family_sums <- sample_sums(ps_family_rel)
cat("Family-level sample sums (should all be 1.0):\n")
print(summary(family_sums))
cat("\nSamples NOT summing to 1.0:\n")
print(family_sums[abs(family_sums - 1) > 0.01])

# Check if ps_genus_rel samples sum to 1
genus_sums <- sample_sums(ps_genus_rel)
cat("\nGenus-level sample sums (should all be 1.0):\n")
print(summary(genus_sums))
cat("\nSamples NOT summing to 1.0:\n")
print(genus_sums[abs(genus_sums - 1) > 0.01])
```

```{r debug-one-plot, eval=FALSE}
# DEBUG: Use this to diagnose individual plot issues
test_ps <- prune_samples(sample_data(ps_family_rel)$Method_Category == "Host-DNA depletion", ps_family_rel)

plot_df <- psmelt(test_ps)
if ("sample_Sample" %in% colnames(plot_df) && !"Sample" %in% colnames(plot_df)) {
  plot_df$Sample <- plot_df$sample_Sample
}

plot_df$Taxon <- plot_df$Family
plot_df$Taxon[is.na(plot_df$Taxon) | plot_df$Taxon == ""] <- "Unclassified"

top_taxa <- plot_df %>%
  dplyr::filter(Taxon != "Unclassified") %>%
  dplyr::group_by(Taxon) %>%
  dplyr::summarise(Total = sum(Abundance), .groups = "drop") %>%
  dplyr::arrange(desc(Total)) %>%
  dplyr::slice_head(n = 12) %>%
  dplyr::pull(Taxon)

plot_df <- plot_df %>%
  dplyr::mutate(Taxon_Display = dplyr::case_when(
    Taxon == "Unclassified" ~ "Unclassified",
    Taxon %in% top_taxa ~ Taxon,
    TRUE ~ "Other"
  ))

plot_df_agg <- plot_df %>%
  dplyr::group_by(Sample, Taxon_Display) %>%
  dplyr::summarise(
    Abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

cat("BEFORE normalization:\n")
totals_before <- plot_df_agg %>%
  dplyr::group_by(Sample) %>%
  dplyr::summarise(total = sum(Abundance), .groups = "drop")
print(totals_before)

plot_df_agg <- normalize_composition_by_sample(plot_df_agg, "Sample", "Abundance")

cat("\nAFTER normalization:\n")
totals_after <- plot_df_agg %>%
  dplyr::group_by(Sample) %>%
  dplyr::summarise(total = sum(Abundance), .groups = "drop")
print(totals_after)

cat("\nAny totals NOT 1.0?\n")
print(totals_after %>% filter(abs(total - 1) > 0.001))

cat("\nTaxa present:\n")
print(table(plot_df_agg$Taxon_Display))
```

---

# Create Visualizations

## Host-DNA Depletion Comparison

```{r host-dna-depletion-family, fig.width=16, fig.height=11}
create_faceted_plot(
  ps_family_rel,
  sample_filter = "Host-DNA depletion",
  title = "Effect of Host-DNA Depletion on Bacterial Composition\n(Top 12 Families)",
  output_file = "final_plots/01_host_dna_depletion_family.png",
  taxon_rank = "Family",
  top_n = 12
)
```

```{r host-dna-depletion-genus, fig.width=16, fig.height=11}
create_faceted_plot(
  ps_genus_rel,
  sample_filter = "Host-DNA depletion",
  title = "Effect of Host-DNA Depletion on Bacterial Composition\n(Top 12 Genera)",
  output_file = "final_plots/02_host_dna_depletion_genus.png",
  taxon_rank = "Genus",
  top_n = 12
)
```

## Filtration Testing

```{r filtration-family, fig.width=16, fig.height=11}
create_faceted_plot(
  ps_family_rel,
  sample_filter = "Filtration testing for Host-DNA depletion",
  title = "Effect of Filtering and host-DNA Depletion on Bacterial Composition\n(Top 12 Families)",
  output_file = "final_plots/03_filtration_testing_family.png",
  taxon_rank = "Family",
  top_n = 12
)
```

```{r filtration-genus, fig.width=16, fig.height=11}
create_faceted_plot(
  ps_genus_rel,
  sample_filter = "Filtration testing for Host-DNA depletion",
  title = "Effect of Filtering and host-DNA Depletion on Bacterial Composition\n(Top 12 Genera)",
  output_file = "final_plots/04_filtration_testing_genus.png",
  taxon_rank = "Genus",
  top_n = 12
)
```

## Enrichment Testing + Host-DNA Depletion

```{r enriched-depletion-family, fig.width=16, fig.height=11}
create_faceted_plot(
  ps_family_rel,
  sample_filter = "Host-DNA depletion - enriched samples",
  title = "Effect of Host-DNA Depletion on Enriched Samples\n(Top 12 Families)",
  output_file = "final_plots/05_enriched_depletion_family.png",
  taxon_rank = "Family",
  top_n = 12
)
```

```{r enriched-depletion-genus, fig.width=16, fig.height=11}
create_faceted_plot(
  ps_genus_rel,
  sample_filter = "Host-DNA depletion - enriched samples",
  title = "Effect of Host-DNA Depletion on Enriched Samples\n(Top 12 Genera)",
  output_file = "final_plots/06_enriched_depletion_genus.png",
  taxon_rank = "Genus",
  top_n = 12
)
```

## Enrichment Testing - Temporal Dynamics

```{r enrichment-temporal-family, fig.width=16, fig.height=12}
create_enrichment_plot_4broths(
  ps_family_rel,
  sample_filter = "Enrichment Testing",
  title = "Enrichment Testing - Temporal Dynamics by Broth (Top 12 Families)",
  output_file = "final_plots/07_enrichment_temporal_family.png",
  taxon_rank = "Family",
  top_n = 12
)
```

```{r enrichment-temporal-genus, fig.width=16, fig.height=12}
create_enrichment_plot_4broths(
  ps_genus_rel,
  sample_filter = "Enrichment Testing",
  title = "Enrichment Testing - Temporal Dynamics by Broth (Top 12 Genera)",
  output_file = "final_plots/08_enrichment_temporal_genus.png",
  taxon_rank = "Genus",
  top_n = 12
)
```

## DNA Extraction Kit Comparison

```{r extraction-kits, fig.width=12, fig.height=11}
create_simple_plot(
  ps_family_rel,
  sample_filter = "DNA extraction kit testing",
  title = "DNA Extraction Kit Comparison\n(Top 12 Families)",
  output_file = "final_plots/09_extraction_kit_family.png",
  taxon_rank = "Family",
  top_n = 12,
  width = 12,
  height = 11
)
```

## Heatmaps

```{r heatmap-all-family, fig.width=18, fig.height=10}
create_heatmap(
  ps_family_rel,
  sample_filter = NULL,
  title = "All Samples - Microbial Diversity (Top 12 Families)",
  output_file = "final_plots/10_all_samples_heatmap_family.png",
  taxon_rank = "Family",
  top_n = 12
)
```

```{r heatmap-all-genus, fig.width=18, fig.height=10}
create_heatmap(
  ps_genus_rel,
  sample_filter = NULL,
  title = "All Samples - Microbial Diversity (Top 12 Genera)",
  output_file = "final_plots/11_all_samples_heatmap_genus.png",
  taxon_rank = "Genus",
  top_n = 12
)
```

```{r heatmap-depletion, fig.width=18, fig.height=10}
create_heatmap(
  ps_family_rel,
  sample_filter = c("Host-DNA depletion", "Filtration testing for Host-DNA depletion", 
                    "Host-DNA depletion - enriched samples"),
  title = "Host-DNA Depletion Methods Comparison (Top 12 Families)",
  output_file = "final_plots/12_host_depletion_heatmap_family.png",
  taxon_rank = "Family",
  top_n = 12
)
```

---

# Read Tracking and Distribution

## Read Tracking

```{r read-tracking, fig.width=14, fig.height=8}
track_long <- data.frame(
  Sample = sample_names(ps),
  input = sample_data(ps)$input,
  filtered = sample_data(ps)$filtered,
  merged = sample_data(ps)$merged,
  nonchim = sample_data(ps)$nonchim,
  Method_Category = sample_data(ps)$Method_Category
) %>%
  pivot_longer(cols = c(input, filtered, merged, nonchim),
               names_to = "Stage", values_to = "Reads") %>%
  mutate(Stage = factor(Stage, levels = c("input", "filtered", "merged", "nonchim")))

p_tracking <- ggplot(track_long, aes(x = Stage, y = Reads, color = Method_Category, group = Sample)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.5, size = 1) +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = category_colors) +
  labs(x = "Pipeline Stage", y = "Read Count",
       title = "Read Tracking Through DADA2 Pipeline",
       color = "Sample Category") +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

if (!dir.exists("final_plots")) dir.create("final_plots", recursive = TRUE)
png("final_plots/13_read_tracking.png", width = 14, height = 8, units = "in", res = 300)
print(p_tracking)
dev.off()

print(p_tracking)
```

## Read Distribution

```{r read-distribution, fig.width=12, fig.height=8}
track_summary <- data.frame(
  Sample = sample_names(ps),
  nonchim = sample_data(ps)$nonchim,
  Method_Category = sample_data(ps)$Method_Category
)

p_distribution <- ggplot(track_summary, aes(x = Method_Category, y = nonchim)) +
  geom_boxplot(aes(fill = Method_Category), alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 2) +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = category_colors) +
  labs(x = "", y = "Non-chimeric Reads",
       title = "Final Read Count Distribution by Category") +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    panel.grid.minor = element_blank()
  )

png("final_plots/14_read_distribution.png", width = 12, height = 8, units = "in", res = 300)
print(p_distribution)
dev.off()

print(p_distribution)
```

---

# Summary Statistics

```{r summary-statistics}
summary_stats <- data.frame(
  Sample = sample_names(ps),
  input = sample_data(ps)$input,
  filtered = sample_data(ps)$filtered,
  merged = sample_data(ps)$merged,
  nonchim = sample_data(ps)$nonchim,
  Method_Category = sample_data(ps)$Method_Category,
  DNA_Conc = sample_data(ps)$DNA_Conc
) %>%
  group_by(Method_Category) %>%
  summarise(
    N_Samples = n(),
    Mean_Input = round(mean(input, na.rm = TRUE), 0),
    SD_Input = round(sd(input, na.rm = TRUE), 0),
    Mean_Filtered = round(mean(filtered, na.rm = TRUE), 0),
    Mean_Merged = round(mean(merged, na.rm = TRUE), 0),
    Mean_Nonchim = round(mean(nonchim, na.rm = TRUE), 0),
    SD_Nonchim = round(sd(nonchim, na.rm = TRUE), 0),
    Min_Nonchim = min(nonchim, na.rm = TRUE),
    Max_Nonchim = max(nonchim, na.rm = TRUE),
    Pct_Retained = round(mean(nonchim / input * 100, na.rm = TRUE), 1),
    Mean_DNA_Conc = round(mean(DNA_Conc, na.rm = TRUE), 1),
    .groups = "drop"
  )

knitr::kable(
  summary_stats,
  caption = "Summary Statistics by Sample Category",
  format.args = list(big.mark = ",")
)

write.csv(summary_stats, "final_plots/read_count_summary.csv", row.names = FALSE)
cat("✓ Saved summary statistics\n")
```

---

# Save R Objects

```{r save-objects}
saveRDS(ps, "final_plots/ps_with_sample_data.rds")
saveRDS(ps_family_rel, "final_plots/ps_family_rel.rds")
saveRDS(ps_genus_rel, "final_plots/ps_genus_rel.rds")
saveRDS(track_df, "final_plots/track_df.rds")

cat("✓ Saved R objects\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```

---

# Files Created

```{r list-files}
png_files <- list.files("final_plots", pattern = "\\.png$", full.names = FALSE)
csv_files <- list.files("final_plots", pattern = "\\.csv$", full.names = FALSE)
rds_files <- list.files("final_plots", pattern = "\\.rds$", full.names = FALSE)

cat("PNG Visualizations (", length(png_files), "):\n", sep = "")
for (file in png_files) cat("  •", file, "\n")

cat("\nData Files (", length(csv_files), "):\n", sep = "")
for (file in csv_files) cat("  •", file, "\n")

cat("\nR Objects (", length(rds_files), "):\n", sep = "")
for (file in rds_files) cat("  •", file, "\n")
```